Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-31 manuscript/manuscriptRW.Rmd
remove word counting from manuscript
update protocol reference in the manuscript

    Word count: `r as.integer(sub("(\\d+).+$", "\\1", system(sprintf("wc -w %s", knitr::current_input()), intern = TRUE))) - 20`

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol (cf.
Supplementary Material 1, henceforth "SM") with instructions how to
perform the simulations, providing input variables on climate and soil
conditions, and the expected output variables. Neither were further site
information (except for the data specified below) nor any data on tree
ingrowth or forest stand features provided prior to the simulation. That
is, the models were run in "blind flight" mode.

@@

The simulations were run in the absence of natural disturbances. Only
the model LPJ-GUESS had to include a background disturbance to increase
chances of shade-intolerant species to establish. The simulations were
set up to sample species-specific ingrowth rates per decade and per ha
in the equilibrium state of the model, typically entailing a "spin-up"
run (as we did not provide any forest data). The modeling teams decided
on the simulated area and how they derived these samples. The exact
length of the simulation was also decided by the modeling teams (Table
\@ref(tab:tableModels)). Further details on how each modeling team
prepared the simulations and the outputs are available in SM 2.

@@

The simulation results were analyzed regarding (1) ingrowth levels
(i.e., ingrowth number per unit time and space), (2) species diversity
of the ingrowth, and (3) ingrowth mortality. We evaluated the species
diversity across models and in relation to the observed data by
calculating the Shannon index based on the relative proportion of the
species in terms of basal area. The Shannon index was calculated for the
establishment ($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the
stand level ($H_{S_{n,s}}$)(Equation \ref{eqpiShannon}). The higher the
value of the index, the higher the species diversity at a particular
site and sample.

@@

## Regeneration niche

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021, they found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic !!! Second, the data were collected for ingrowth above a 7
cm threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This, may at
least parlty explain e.g. the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

Therefore, we recommend that more effort should be invested into
collecting harmonised datasets on tree establishment. We emphasize that
datasets like EuFoRIa used here are invaluable and should be further
expanded in both their spatial extent (e.g., towards boreal and
Mediterranean conditions) as well as in time (e.g., continuing the
monitoring into the future). Such data will allow for a better
evaluation of forest models and help to reduce the uncertainty in their
projections, which is crucial when they are used as tools for predicting
e.g. the impacts of anthropogenic climate change.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available here:( link to be added). The code and data that support the
findings of this study are openly available at zenodo and gihub (links
to be added).

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project XXXX . KB received support from the European
Research Council under the European Union's Horizon 2020 research and
innovation program (Grant Agreement 101001905). JB acknowledges funding
from the long-term research development project no. RVO 67985939 from
the Czech Academy of Sciences. MM acknowledges financial support from
I-Maestro (Innovative forest management strategies for a resilient
bioeconomy under climate change and disturbances, grant no. 22035418,
2019--2022) funded by the ERA-NET Cofund ForestValue. KM received
funding from the Project No. CZ.02.1.01/0.0/0.0/16_019/0000803 financed
by OP RDE, grant no. ITMS2014+313011W580 supported by the Integrated
Infrastructure Operational Programme funded by the ERDF.

>>>>

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol [@díaz-yáñez]
with instructions how to perform the simulations, providing input
variables on climate and soil conditions, and the expected output
variables. Neither were further site information (except for the data
specified below) nor any data on tree ingrowth or forest stand features
provided prior to the simulation. That is, the models were run in "blind
flight" mode.

@@

The simulations were run in the absence of natural disturbances. Only
the model LPJ-GUESS had to include a background disturbance to increase
chances of shade-intolerant species to establish. The simulations were
set up to sample species-specific ingrowth rates per decade and per ha
in the equilibrium state of the model, typically entailing a "spin-up"
run (as we did not provide any forest data). The modeling teams decided
on the simulated area and how they derived these samples. The exact
length of the simulation was also decided by the modeling teams (Table
\@ref(tab:tableModels)). Further details on how each modeling team
prepared the simulations and the outputs are available in Supplementary
Material 1.

@@

The simulation results were analyzed regarding (1) ingrowth levels
(i.e., ingrowth number per unit time and space), (2) species diversity
of the ingrowth, (3) ingrowth mortality, (4) model performance and model
traits, and (5) ingrowth gradients on regeneration niches. We evaluated
the species diversity across models and in relation to the observed data
by calculating the Shannon index based on the relative proportion of the
species in terms of basal area. The Shannon index was calculated for the
establishment ($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the
stand level ($H_{S_{n,s}}$) (Equation \ref{eqpiShannon}). The higher the
value of the index, the higher the species diversity at a particular
site and sample.

@@

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021, they found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This, may at
least partly explain e.g. the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

Therefore, we recommend that more effort should be invested into
collecting harmonized datasets on tree establishment. We emphasize that
datasets like EuFoRIa used here are invaluable and should be further
expanded in both their spatial extent (e.g., towards boreal and
Mediterranean conditions) as well as in time (e.g., continuing the
monitoring into the future). Such data will allow for a better
evaluation of forest models and help to reduce the uncertainty in their
projections, which is crucial when they are used as tools for predicting
e.g. the impacts of anthropogenic climate change.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available here @díaz-yáñez . The code and data that support the findings
of this study are openly available at zenodo (*link to be added*) and
gihub (*link to be added*).

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project 200020_188882 / 1 . KB received support from
the European Research Council under the European Union's Horizon 2020
research and innovation program (Grant Agreement 101001905). JB
acknowledges funding from the long-term research development project no.
RVO 67985939 from the Czech Academy of Sciences. MM acknowledges
financial support from I-Maestro (Innovative forest management
strategies for a resilient bioeconomy under climate change and
disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. KM received funding from the Project No.
CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant no.
ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF.
